\chapter{Replication report}
\label{chapter2}

\section{Anomaly Detection Replication}

This section presents a replication and analysis of a lightweight anomaly detection approach proposed for edge-based Kubernetes clusters. The goal is not only to reproduce reported behaviour, but also to evaluate its robustness under realistic edge preprocessing conditions.

\subsection{Introduction}

Edge Kubernetes environments demand monitoring solutions that are both effective and lightweight. Since edge computing exists to achieve low latency and high bandwidth, any added computation for monitoring must be carefully optimised to operate within resource constraints, intermittent connectivity, geographic distribution, and the need for autonomous operation \cite{rasouli2025false}.

Anomaly detection supports lightweight monitoring by enabling real-time decisions about node health. By identifying abnormal behaviour locally, edge nodes can selectively transfer only relevant data, reducing network load and storage cost. Additional benefits include improved security posture, predictive maintenance capabilities, context-aware resource management, and greater system autonomy \cite{fritzsche2024lightweight}.

For anomaly detection in this context, K-means and the Negative Selection Algorithm (NSA) offer complementary strengths. K-means provides computationally simple clustering-based analysis on large unlabelled datasets, while NSA offers lightweight multi-dimensional classification that is well-suited to the diverse operating conditions of edge environments \cite{fritzsche2024lightweight}.

\subsection{Method Overview}

Fritzsche et al.\ demonstrated that a combination of K-means and NSA detection achieves effective anomaly detection in edge Kubernetes clusters \cite{fritzsche2024lightweight}. This section replicates their approach under controlled conditions to evaluate detector behaviour and sensitivity to preprocessing.

The NSA algorithm takes as input three sets: a \textit{self} set representing normal behaviour, a \textit{monitor} set of observations to classify, and a \textit{detector} set generated during training. Given a maximum detector count $n$, detectors are randomly generated in feature space and retained only if they do not match any self-region data point. At runtime, each observation in the monitor set is classified as \textit{self} or \textit{non-self} based on proximity to detectors.

NSA is well-suited to edge environments because it requires only normal-behaviour training data (one-class learning), imposes minimal computational overhead at inference time, and naturally adapts to multi-dimensional feature spaces without requiring labelled anomaly examples.

\subsection{Experimental Setup (offline)}

An offline evaluation was conducted to isolate detector behaviour from runtime cluster effects, enabling controlled algorithm evaluation and sensitivity analysis independent of network variability or collection artefacts.

\subsubsection{Metrics}

Synthetic edge metrics were generated using normal distributions to simulate CPU, memory, and network utilisation under typical operating conditions. Two modes of anomaly injection were used---subtle (transient point anomalies) and sustained (prolonged multi-dimensional deviations)---which present varying sensitivity to preprocessing.

Subtle point anomalies were designed to emulate transient workload disturbances common in edge environments. However, after rolling-window smoothing and feature normalisation, these anomalies became geometrically embedded within the normal data distribution. As a result, no distinct non-self region emerged, rendering NSA incapable of detecting such deviations by design.

In contrast, sustained multi-dimensional anomalies introduce prolonged deviations in system behaviour, forming a distinct abnormal operating regime. When excessive smoothing is avoided, this regime remains geometrically separable from normal workloads, allowing NSA to successfully identify non-self regions.

\subsubsection{Preprocessing}

Preprocessing was completed to achieve edge realism. Edge monitoring systems must trade precision for stability, bandwidth efficiency, and robustness.

Rolling-window smoothing was applied to each resource metric to reflect realistic edge telemetry behaviour, where lightweight monitoring systems aggregate measurements over short intervals to suppress noise and reduce communication overhead. This approach stabilises metric trends while preserving sustained workload changes relevant for scheduling decisions.

Feature normalisation was applied to ensure scale invariance across heterogeneous resource metrics, enabling fair distance-based comparison between CPU, memory, and network utilisation. This reflects standard practice in ML-driven resource management pipelines and prevents dominance by high-magnitude features.

While such preprocessing improves robustness and reflects realistic edge monitoring constraints, it also reduces the geometric separability of subtle anomalies, highlighting a fundamental trade-off between noise suppression and anomaly detectability.

\subsubsection{NSA Configuration}

All NSA parameters were held constant across experiments to ensure fair comparison.

\begin{table}[h]
    \centering
    \begin{tabular}{ll}
        \hline
        \textbf{Parameter} & \textbf{Value} \\
        \hline
        Detector count & 200 \\
        Detector radius & 1.5 \\
        Feature space & 3-dimensional (CPU, memory, network) \\
        Training type & Normal-only (one-class) \\
        \hline
    \end{tabular}
    \caption{NSA configuration parameters for offline evaluation.}
    \label{tab:nsa-params}
\end{table}

\subsection{Evaluation Methodology}

Point-wise detection was used to discover individual spikes, which is suitable for scheduler evaluation since it gives high granularity and precision, resulting in the exact identification of decision successes and failures. Precision and recall metrics provided a balanced evaluation; measuring both the accuracy of the detected nodes and the coverage of total positive nodes found.  

The results were interpreted by calculating the distance from the normal of each node that was detected to decide whether it was located in the self or non-self region. If a large proportion of detected nodes were located in the self region, this signaled over sensitivity to the preprocessing pipeline.

\subsection{Results and Analysis}

\subsubsection{Dataset A: Subtle Anomalies}


\begin{table}[h]
    \centering
    \begin{tabular}{ll}
        \hline
        \textbf{Metric} & \textbf{Value} \\
        \hline
        Distance from normal & 0.25 \\
        Precision & 0.00 \\
        Recall & 1.00 \\
        \hline
    \end{tabular}
    \caption{NSA results on Dataset A: subtle point anomalies.}
    \label{tab:results-subtle}
\end{table}

The subtle anomalies were geometrically embedded within the normal distribution after preprocessing, leaving no distinct non-self region for NSA detectors to cover. While recall reached 1.0 (all points were flagged), precision was 0.0, indicating that the detector could not distinguish anomalous points from normal behaviour.

\subsubsection{Dataset B: Sustained Anomalies (default window size)}

\begin{table}[h]
    \centering
    \begin{tabular}{ll}
        \hline
        \textbf{Metric} & \textbf{Value} \\
        \hline
        Distance from normal & 0.34 \\
        Precision & 0.00 \\
        Recall & 1.00 \\
        \hline
    \end{tabular}
    \caption{NSA results on Dataset B with default window size.}
    \label{tab:results-sustained-default}
\end{table}

Despite the anomalies being stronger and more prolonged than in Dataset A, excessive smoothing from the default rolling window collapsed their separability. The distance from normal (0.34) was marginally higher than Dataset A but still insufficient for NSA to establish a meaningful non-self boundary.

\subsubsection{Dataset B: Sustained Anomalies (window size = 2)}

\begin{table}[h]
    \centering
    \begin{tabular}{ll}
        \hline
        \textbf{Metric} & \textbf{Value} \\
        \hline
        Distance from normal & 1.93 \\
        Precision & 0.48 \\
        Recall & 1.00 \\
        \hline
    \end{tabular}
    \caption{NSA results on Dataset B with reduced window size (w=2).}
    \label{tab:results-sustained-w2}
\end{table}

With the reduced window size, the distance from normal increased substantially to 1.93, confirming that a geometrically distinct non-self region emerged. NSA achieved precision of 0.48 while maintaining perfect recall, indicating that the detector successfully identified all anomalous periods albeit with some false positives during transitional phases. This result validates that NSA performs as expected when its core assumption---geometric separability between normal and anomalous behaviour---is preserved through appropriate preprocessing choices.
